Legion: Expressing Locality and Independence with Logical Regions

### 2012-04-27 doc/sc2012/circuit.tex
edits

In this section we give an overview of Legion using an example.
Listing~\ref{lst:code_ex} shows code for an electrical
circuit simulation, which illustrates the core features of the programming model.

The circuit simulation takes a collection of
of wires and nodes where wires meet.  
The simulation repeatedly updates
currents, distributes charges, and updates voltages for as many
time steps as the simulation demands.  We first
describe the partitioning of regions in Section~\ref{subsec:partitioning}
and then describe the data type declarations in Section~\ref{subsec:datatypes}.

@@

We partition the shared nodes a second time using a
different coloring into the sets of ghost nodes for each task.  This 
partition is noteworthy because the subregions are {\em not} disjoint:
ghost nodes may be read by multiple tasks, and thus a node $n$  may belong 
to more than one ghost region, which corresponds to $n$ being assigned
more than one color by the coloring {\tt node\_neighbor\_map}.
One of the ghost subregions is shown in Figure~\ref{sfig:part_fig:g_i}.

@@

\subsection{Data Types}
\label{subsec:datatypes}
The two basic data types in the circuit simulation are {\tt Node}s and
{\tt Wire}s declared on lines 1 and 2 of Listing~\ref{lst:code_ex}.
As is standard in region-based systems, allocating a {\tt Node}, {\tt
  Wire}, or any value in the heap requires naming the logical region
in which the value is allocated.  Furthermore, pointers to heap values
declare the region to which they point.  For example, the definition
of {\tt Wire} is parameterized on the region {\tt rn} to which the
{\tt Node} pointers in fields {\tt in\_nodes} and {\tt out\_nodes}
point, which guarantees that the two endpoints of a wire are always in
the same region.

>>>>

Listing~\ref{lst:code_ex} shows code for an electrical
circuit simulation, which illustrates the core features of the Legion programming model.
The circuit simulation takes a collection of
of wires and nodes where wires meet.  
The simulation repeatedly updates
currents, distributes charges, and updates voltages for as many
time steps as the simulation demands.  We first
describe the partitioning of regions in Section~\ref{subsec:partitioning}
and then describe the data type declarations in Section~\ref{subsec:datatypes}.

@@

We partition the shared nodes a second time using a
different coloring into the sets of ghost nodes for each task.  This 
partition is noteworthy because the subregions are not disjoint:
ghost nodes may be read by multiple tasks, and thus a node $n$  may belong 
to more than one ghost region, which corresponds to $n$ being assigned
more than one color by the coloring {\tt node\_neighbor\_map}.
One of the ghost subregions is shown in Figure~\ref{sfig:part_fig:g_i}.

@@

\subsection{Data Types}
\label{subsec:datatypes}
The two basic data types in the circuit simulation are {\tt Node}s and
{\tt Wire}s declared on lines 1 and 2 of Listing~\ref{lst:code_ex}.
As is standard in region-based systems, allocating a {\tt Node}, {\tt
  Wire}, or any value in the heap requires naming the logical region
in which the value is allocated.  For example, {\tt new Node@r} returns
a reference to a new value of {\tt Node} type in region {\tt r}.  
Listing~\ref{lst:code_ex} illustrates one way of constructing partitioned
data structures in Legion: populate a region with data (using {\tt new}) and then partition it
into subregions.  Legion also supports partitioning an empty region into subregions
and then allocating data in the subregions; we use both approaches in our
application discussed in Section~\ref{sec:ex}.

Pointers to region values declare the region to which they point.  For
example, the definition of {\tt Wire} is parameterized on the region
{\tt rn} to which the {\tt Node} pointers in fields {\tt in\_nodes}
and {\tt out\_nodes} point, which guarantees that the two endpoints of
a wire are always in the same region.

### 2012-04-27 doc/sc2012/legion.tex
edits

Modern parallel machines are increasingly complex, with deep,
distributed memory hierarchies and heterogeneous processing units.  On
these architectures, it is crucial for performance that the programmer
and the compiler be able to reason about {\em locality} (data resident
close to computation that uses it) and {\em independence} (computations
operating on disjoint data).  Most contemporary
languages have no facilities for the programmer to express locality
and independence.  The few languages that do focus primarily on
array-based locality \cite{Fatahalian06,CHAPEL04,UPC99} and 
avoid irregular pointer data structures.
%encode such structures in arrays, which 
%data structures encoded in arrays which are impossible for compilers
%and runtime systems to reason about with regard to locality and independence.

In this paper we describe Legion, a parallel programming system based
on using {\em logical regions} to describe the organization of data.
A logical region names a set of objects.  Logical regions are first
class values in Legion, and in particular may be passed as arguments
to distinguished functions called {\em tasks} that access the data in
those regions, providing locality information.  Logical regions may be
{\em partitioned} into (usually) disjoint {\em subregions}, providing
information for determining independence of computations.
Furthermore, computations access logical regions with particular {\em
  privileges} (e.g., {\em read-only} and {\em read-write}, among
others) and {\em coherence} (e.g., {\em exclusive access} and {\em
  atomic access}, among others).  Priveleges express how a task may
use its region arguments, providing data dependence information that
is used to guide the extraction of parallelism (e.g., if two tasks
access the region $r$with {\em read-only} priveleges the two tasks
can potentially be run in parallel, at least with respect to $r$),
and coherence properties express the required semantics of concurrent
accesses to the region.  For example, if tasks $x$ and $y$ access
region $r$ with {\em exclusive} access, then the effects of $x$ and
$y$ on $r$ are guaranteed to be the same as if they had executed in
the sequential order expressed by the program; on the other hand, if
the tasks access $r$ with {\em atomic} coherence, then Legion
guarantees only atomicity of the tasks with respect to $r$: either $x$
appears to run entirely before $y$ or vice versa.

To introduce the programming model, we begin with an extended example
of a circuit simulation programmed in Legion in
Section~\ref{sec:ex}. This example illustrates regions, tasks,
permissions and coherence properties, the interactions between them,
and how these building blocks are assembled into a Legion
program.  Subsequent sections each describe a contribution in the
implementation and evaluation of Legion:

@@

\item A crucial part of orchestrating a program's execution is how
tasks are assigned (or {\em mapped}) to processors and how regions are
mapped to specific memory units within the machine.  Often there is
application-specific information that, if exploited, results in better
mappings than a generic, one-size-fits-all mapping strategy.  We describe
a {\em mapping interface} that allows programmers to give the SOOP a specification
of how to map tasks and regions for a specific application, or even part of an
application.  This mapping API is carefully defined so that
any user-supplied mapping strategy can only affect the performance of applications,
not their correctness.  Thus, user mappings only affect the time and resources required
to run the program and cannot change the program's output (Section~\ref{sec:mapping}).

@@

\end{itemize}
\input{code_ex}
\input{circuit}
\input{part_fig}
\input{soop}
\input{mapping}
\input{experiments}
\input{related}
\input{conclusion}

>>>>

Modern parallel machines are increasingly complex, with deep,
distributed memory hierarchies and heterogeneous processing units.  Because the
costs of communication within these architectures vary by several orders of magnitude, 
the penalty for mistakes in the placement of data or computation is very poor performance.  Thus,
to achieve high performance it is crucial that the programmer
and the compiler be able to reason about {\em locality} (data resident
close to computation that uses it) and {\em independence} (computations
operating on disjoint data, and therefore not requiring communication and able to be placed in possibly distant parts of the machine).  Most contemporary
programming systems have no facilities for the programmer to express locality
and independence.  The few languages that do focus primarily on
array-based locality \cite{Fatahalian06,CHAPEL04,UPC99} and 
avoid irregular pointer data structures.
%encode such structures in arrays, which 
%data structures encoded in arrays which are impossible for compilers
%and runtime systems to reason about with regard to locality and independence.

In this paper we describe Legion, a parallel programming system based
on using {\em logical regions} to describe the organization of data.
A logical region names a set of objects.  Logical regions are first
class values in Legion, and in particular may be passed as arguments
to distinguished functions called {\em tasks} that access the data in
those regions, providing locality information.  Logical regions may be
{\em partitioned} into (usually) disjoint {\em subregions}, providing
information for determining independence of computations.
Furthermore, computations access logical regions with particular {\em
  privileges} (e.g., {\em read-only} and {\em read-write}, among
others) and {\em coherence} (e.g., {\em exclusive access} and {\em
  atomic access}, among others).  Priveleges express how a task may use
its region arguments, providing data dependence information that
is used to guide the extraction of parallelism.  For example, if two tasks
access the same region with {\em read-only} privileges the two tasks
can potentially be run in parallel.
Coherence properties express the required semantics of concurrent
accesses to the region.  For example, if tasks $x$ and $y$ access
region $r$ with {\em exclusive} access, then Legion guarantees that the effects of $x$ and
$y$ on $r$ are guaranteed to be the same as if they had executed in
the sequential order expressed by the program. On the other hand, if
the tasks access $r$ with {\em atomic} coherence, then Legion
guarantees only atomicity of the tasks with respect to $r$: either $x$
appears to run entirely before $y$ or vice versa.

Legion programs are written using logical regions, each of which
has one or more {\em physical instances} when the program executes.  Often
logical regions correspond to a single physical instance, but there are common situations
in which it is desirable to have multiple physical copies of a logical region's
contents in different parts of the machine (e.g., to duplicate and distribute read-only
data) or even to have physical instances of a logical region that temporarily diverge
in their contents (e.g., to keep an old copy of a region to break a write-after-read dependency, or to 
allow independent, local reductions that are later combined).

To introduce the programming model, we begin with an extended example
of a circuit simulation in
Section~\ref{sec:ex}. This example illustrates regions, tasks,
permissions and coherence properties, the interactions between them,
and how these building blocks are assembled into a Legion
program.  Subsequent sections each describe a contribution in the
implementation and evaluation of Legion:

\input{code_ex}

@@

\item A crucial part of orchestrating a program's execution is how
tasks are assigned (or {\em mapped}) to processors and how physical instances of logical regions are
mapped to specific memory units within the machine.  Often there is
application-specific information that, if exploited, results in better
mappings than a generic, one-size-fits-all mapping strategy.  We describe
a {\em mapping interface} that allows programmers to give the SOOP a specification
of how to map tasks and regions for a specific application, or even part of an
application.  This mapping API is carefully defined so that
any user-supplied mapping strategy can only affect the performance of applications,
not their correctness.  Thus, user mappings only affect the time and resources required
to run the program and cannot change the program's output (Section~\ref{sec:mapping}).

@@

\end{itemize}

\input{circuit}
\input{part_fig}
\input{soop}
\input{mapping}
\input{experiments}
\input{related}
\input{conclusion}
